---
title: "Linear Programming"
author: Abhinav Anand
date: 

output:
  pdf_document:
    keep_tex: true

fontsize: 11pt
documentclass: article
geometry: margin = 1.5in

linkcolor: blue
urlcolor: red
citecolor: magenta

citation_package: natbib
bibliography: Working_Paper.bib

header-includes:
   - \linespread{1.25}
   - \usepackage{amsmath}


---

```{r setup, eval=T, message=FALSE, warning=T, include=FALSE}

library(tidyverse)
library(rmarkdown)
library(knitr)
library(xml2)

knitr::opts_chunk$set(echo = T, 
                      warning = T, 
                      message = F, 
                      eval = T, 
                      include = T
                      )
```


# Background

Linear programming involves minimizing linear cost functions subject to linear inequality constraints.[^1]

[^1]: This is without loss of generality since a maximization program is the negative of the corresponding minimization program.

## Illustration:

To illustrate, consider a classic portfolio analysis problem: bonds generate 5% returns, stocks generate 8% returns. The total budget is $1000. How much of each asset should be bought?

We can translate this problem into a linear programming problem:

$$
\text{max } 0.05b+0.08s: b+s\leq 1000, b \geq 0, s \geq 0
$$
where $b, s$ respectively are the amount (in dollars) invested in bonds and stocks.

### The Feasible Set

Any combination of bonds and stocks that satisfies the inequalities: $b ,s \geq 0$ and 
$b+s\leq 1000$ is *feasible*. In the problem above, the feasible set is the following triangular region:

```{r plot_ex_feasible, fig.width=4, fig.height=4}

b <- 0:1000
s <- 1000 - b

ggplot(data.frame(cbind(b, s)), aes(b, s)) +
  geom_line() +
  geom_vline(xintercept = 0) + #vertical line
  geom_hline(yintercept = 0) + #horizontal line
  theme_minimal()

```

In the problem above and more generally, in any linear programming problem, the feasible set is an intersection of *half-spaces*. Clearly, the more constraints we have, the smaller the feasible set is. The feasible set in general can be of three varieties:

1. It is empty. In this case there is no solution.
2. It is not empty but the cost function is unbounded over it. The minimum cost is $-\infty$.
3. It is not empty *and* the cost function is bounded over it.

Only the last case has practical value.

### Finding the Minimum

In principle, to find the minimum, all we need to do is to evaluate the objective function at all feasible points; and then see which point yields the minimum. Clearly, this is not practical since there are a continuum of points in this case. Hence we prefer to reach the minimum in a more systematic way.

