---
title: "Linear Programming"
author: Abhinav Anand
date: 

output:
  pdf_document:
    keep_tex: true

fontsize: 11pt
documentclass: article
geometry: margin = 1.5in

linkcolor: blue
urlcolor: red
citecolor: magenta

citation_package: natbib
bibliography: Working_Paper.bib

header-includes:
   - \linespread{1.25}
   - \usepackage{amsmath}


---

```{r setup, eval=T, message=FALSE, warning=T, include=FALSE}

library(tidyverse)
library(rmarkdown)
library(knitr)
library(xml2)

knitr::opts_chunk$set(echo = T, 
                      warning = T, 
                      message = F, 
                      eval = T, 
                      include = T
                      )
```


# Background

Linear programming involves minimizing linear cost functions subject to linear inequality constraints. 

## Illustration:

To illustrate, consider a classic portfolio analysis problem: bonds generate 5% returns, stocks generate 8% returns. The total budget is $1000. How much of each asset should be bought?

We can translate this problem into a linear programming problem:

$$
\text{max } 5b+8s: b+s\leq 1000, b \geq 0, s \geq 0
$$
where $b, s$ respectively are the amount (in dollars) invested in bonds and stocks.

Any combination of bonds and stocks that satisfies the inequalities: $b ,s \geq 0$ and 
$b+s\leq 1000$ is *feasible*. In the problem above, the feasible region is the following triangular region:

```{r plot_ex_feasible}

b <- 0:1000
s <- 1000 - b

ggplot(data.frame(cbind(b, s)), aes(b, s)) +
  geom_line() +
  geom_vline(xintercept = 0) + #vertical line
  geom_hline(yintercept = 0) #horizontal line

```

